The Revisited Hidden Weight Bit Function

### 2025-03-17 RevisedHiddenWeight.tex
Typos and bibliography

\begin{definition}[Walsh transform, restricted Walsh transform]\label{def:walsh_transform}
	The Walsh transform at $a \in \F_2^n$ of a Boolean function $f\in\BN$ restricted to a subset $S\subseteq \F_2^n$ is defined as $\wt{f,S} (a) = \sum_{x\in S} (-1)^{f(x)+a \cdot x}$. The (unrestricted) Walsh transform of $f$ is then defined as $\wt{f}=\wt{f,\F_2^n}$. For any integer $k$, we also set $\wt{f,k}=\wt{f,\Ekn{k}{n}}$.
\end{definition}

@@

Various weightwise affine functions (\ie, belonging to $\WD{1}{n}$) have been exhibited, such as in~\cite{TOSC:CarMeaRot17} where the bent functions in Propositions $1$ and $2$ are weightwise affine, or in~\cite{DAM:GinMea22} to show than no weightwise perfectly balanced function is weightwise affine for $n\geq 8$. The arguably best known example of weightwise affine function is the Hidden Weight Bit Function introduced in~\cite{IEEE:Bryant91}, the one obtained by fixing $f_0=0$ and $f_k=x_k$ for $k \in [1,n]$. The cryptographic properties of this function have been studied in~\cite{DAM:WCST14}, showing good algebraic properties for this function.

@@

In~\cite{DAM:MeaOza24}, the parameters of different functions from $\WD{1}{n}$ and $\WD{2}{n}$ are studied experimentally for $n$ up to $20$, and lower bounds are given for the nonlinearity for all $n$. These bounds focus on cyclic weightwise quadratic functions and involve sums of binomial coefficients. For simplicity, in the following we provide only the nonlinearity values of the majority function and \hwbf{}, as these bounds will be used for comparison.

@@

Accordingly, we can study the restricted Walsh transform of the revisited \hwbf{} by analyzing the one of the following Boolean function $d_n\in\BN$.

@@

\begin{proposition}\label{prop:recursiveDnka}
Let $n\geq 0$ be an even integer. Then for all integers $k$ and all binary vectors $a\in\F_2^n$, the following hold true:
	\begin{itemize}
		\item We have $\Dnka 00a=1$ and $\Dnka 0ka=0$ if $k\neq 0$.
		\item If $n\geq 2$, then for $b=(a_1,\dots,a_{n-2})\in\F_2^{n-2}$, we have:
		\begin{align*}
			\Dnka{n}{k}{a}&=\Dnka{n-2}{k}{b}+\left((-1)^{a_{n-1}} + (-1)^{a_{n}}\right)\Dnka{n-2}{k-1}{b}\\
			&\qquad+ (-1)^{1+a_{n-1}+a_n} \Dnka{n-2}{k-2}{b}.
		\end{align*}
	\end{itemize}
\end{proposition}

@@

\begin{remark}\label{rmk:dependence_dnka}
    The recursive formula of Proposition~\ref{prop:recursiveDnka} gives different cases depending on the values of the pair $(a_{n-1},a_n)$. However, the same reasoning applies with any pair of the form $(a_{2i-1}, a_{2i})$ for $i\in[1,n/2]$. Therefore, the value of $\Dnka{n}{k}{a}$ depends only on the number of pairs $(a_{2i-1}, a_{2i})$ in $a$ being $(0,0)$, $(1,1)$, $(0,1)$ or $(1,0)$. We give the values of $\Dnka{2}{k}{a}$ for $n=2$ in Table~\ref{tab:Dnka}. These values together with Proposition~\ref{prop:recursiveDnka} are sufficient to determine any $\Dnka{n}{k}{a}$.
\end{remark}

@@

\begin{proof}
    If $p=q$, we have $P_a(z)=(z^4-6z^2+1)^p\cdot(z^2+1)^r$, which can be seen as a polynomial in $z^2$. Thus, the coefficient $\Dnka nka$ in $P_a(z)$ of $z^k$ for odd $k$ must be zero. It follows that also $\Dnka n{n-k}a=0$ by Proposition~\ref{proposition:symmetry}.
\end{proof}

@@

\begin{proof}
	We use the expressions from Table~\ref{tab:Dnka_small_k}, and replace $p$, $q$ and $r$ by their respective values. For instance, for the case $n=16m$, we replace $p$, $q$ and $r$ by $m+s(s-1)$, $m+s(s+1)$ and $6m-2s^2$ in the expression $\Dnka n2a=2(p-q)^2-3(p+q)+r$, which yields $\Dnka n2a=0$.~\footnote{This checking can be automatized by a computer algebra system.
	\ifnum\anon=0
		For this, we refer to the corresponding \textsf{SageMath} implementation, accessible via this link: \url{https://github.com/se-tim/Revisited-HWBF.git}.
	\fi}
\end{proof}

@@

This method is based on the identity given in Proposition~\ref{prop:WT}.
For this, consider a vector $b\in\mathbb F_2^n$; we write $(p,q,r)=(p(b),q(b),r(b))$. Also, for $k\in[1,n]$, let $b_k=b+\pi(e_k)$ and $(p_k,q_k,r_k)=(p(b_k),q(b_k),r(b_k))$, where $\pi:[1,n]\to[1,n]$ is again the permutation sending the first $n/2$ elements to the odd positions and the $n/2$ last ones to the even positions. Notice first that for every integer $k\in[1,n]$, it holds that $(p_k,q_k,r_k)\in\{(p\pm1,q,r\mp 1),(p,q\pm 1,r\mp 1)\}$. Furthermore, observe that if $k\in[0,n/2]$ and $(p_k,q_k,r_k)=(p+\alpha,q+\beta,r+\gamma)$ for $\{\alpha,\beta,\gamma\}=\{0,\pm 1\}$ with $\gamma\neq 0$, then we have $(p_{k+n/2},q_{k+n/2},r_{k+n/2})=(p+\alpha',q+\beta',r+\gamma')$ for some $\{\alpha',\beta',\gamma'\}=\{0,\pm 1\}$ with $\gamma'\neq 0$ which only depend on $\alpha$, $\beta$ and $\gamma$. Explicitly, we have $(\alpha',\beta',\gamma')=(\alpha,\beta,\gamma)$ if $\gamma=1$ and $(\alpha',\beta',\gamma')=(\beta,\alpha,\gamma)$ if $\gamma=-1$.

@@

In Table~\ref{table:comparisonsDegAI}, we also display the algebraic degree of different weightwise quadratic function. We observe that the revisited \hwbf{} is the function with the highest degree, together with the \hwbf{}.

@@

\ifnum\full=0
	\bibliographystyle{splncs04}
	\bibliography{add}
\else
	\bibliographystyle{alpha}
	\bibliography{add}
\fi

>>>>

\begin{definition}[Walsh transform]\label{def:walsh_transform}
	The Walsh transform at $a \in \F_2^n$ of a Boolean function $f\in\BN$ restricted to a subset $S\subseteq \F_2^n$ is defined as $\wt{f,S} (a) = \sum_{x\in S} (-1)^{f(x)+a \cdot x}$. The (unrestricted) Walsh transform of $f$ is then defined as $\wt{f}=\wt{f,\F_2^n}$. For any integer $k$, we also set $\wt{f,k}=\wt{f,\Ekn{k}{n}}$.
\end{definition}

@@

Various weightwise affine functions (\ie, belonging to $\WD{1}{n}$) have been exhibited, such as in~\cite{TOSC:CarMeaRot17} where the bent functions in Propositions $1$ and $2$ are weightwise affine, or in~\cite{DAM:GinMea22} to show that no weightwise perfectly balanced function is weightwise affine for $n\geq 8$. The arguably best known example of weightwise affine function is the Hidden Weight Bit Function introduced in~\cite{IEEE:Bryant91}, the one obtained by fixing $f_0=0$ and $f_k=x_k$ for $k \in [1,n]$. The cryptographic properties of this function have been studied in~\cite{DAM:WCST14}, showing good algebraic properties for this function.

@@

In~\cite{DAM:MeaOza24}, the parameters of different functions from $\WD{1}{n}$ and $\WD{2}{n}$ are studied experimentally for $n$ up to $20$, and lower bounds are given for the nonlinearity for all $n$. These bounds focus on cyclic weightwise quadratic functions and involve sums of binomial coefficients. For simplicity, in the following, we provide only the nonlinearity values of the majority function and \hwbf{}, as these bounds will be used for comparison.

@@

Accordingly, we can study the restricted Walsh transform of the revisited \hwbf{} by analyzing the following Boolean function $d_n\in\BN$.

@@

\begin{proposition}\label{prop:recursiveDnka}
Let $n\geq 0$ be an even integer. Then for all integers $k$ and all binary vectors $a\in\F_2^n$, the following hold:
	\begin{itemize}
		\item We have $\Dnka 00a=1$ and $\Dnka 0ka=0$ if $k\neq 0$.
		\item If $n\geq 2$, then for $b=(a_1,\dots,a_{n-2})\in\F_2^{n-2}$, we have:
		\begin{align*}
			\Dnka{n}{k}{a}&=\Dnka{n-2}{k}{b}+\left((-1)^{a_{n-1}} + (-1)^{a_{n}}\right)\Dnka{n-2}{k-1}{b}\\
			&\qquad+ (-1)^{1+a_{n-1}+a_n} \Dnka{n-2}{k-2}{b}.
		\end{align*}
	\end{itemize}
\end{proposition}

@@

\begin{remark}\label{rmk:dependence_dnka}
    The recursive formula of Proposition~\ref{prop:recursiveDnka} gives different cases depending on the values of the pair $(a_{n-1},a_n)$. However, the same reasoning applies to any pair of the form $(a_{2i-1}, a_{2i})$ for $i\in[1,n/2]$. Therefore, the value of $\Dnka{n}{k}{a}$ depends only on the number of pairs $(a_{2i-1}, a_{2i})$ in $a$ being $(0,0)$, $(1,1)$, $(0,1)$ or $(1,0)$. We give the values of $\Dnka{2}{k}{a}$ for $n=2$ in Table~\ref{tab:Dnka}. These values together with Proposition~\ref{prop:recursiveDnka} are sufficient to determine any $\Dnka{n}{k}{a}$.
\end{remark}

@@

\begin{proof}
    If $p=q$, we have $P_a(z)=(z^4-6z^2+1)^p\cdot(z^2+1)^r$, which can be seen as a polynomial in $z^2$. Thus, the coefficient $\Dnka nka$ in $P_a(z)$ of $z^k$ for odd $k$ must be zero. It also follows that $\Dnka n{n-k}a=0$ by Proposition~\ref{proposition:symmetry}.
\end{proof}

@@

\begin{proof}
	We use the expressions from Table~\ref{tab:Dnka_small_k}, and replace $p$, $q$ and $r$ by their respective values. For instance, for the case $n=16m$, we replace $p$, $q$ and $r$ by $m+s(s-1)$, $m+s(s+1)$ and $6m-2s^2$, respectively, in the expression $\Dnka n2a=2(p-q)^2-3(p+q)+r$, which yields $\Dnka n2a=0$.~\footnote{This checking can be automatized by a computer algebra system.
	\ifnum\anon=0
		For this, we refer to the corresponding \textsf{SageMath} implementation, accessible via this link: \url{https://github.com/se-tim/Revisited-HWBF.git}.
	\fi}
\end{proof}

@@

This method is based on the identity given in Proposition~\ref{prop:WT}.
For this, consider a vector $b\in\mathbb F_2^n$; we write $(p,q,r)=(p(b),q(b),r(b))$. Also, for $k\in[1,n]$, let $b_k=b+\pi(e_k)$ and $(p_k,q_k,r_k)=(p(b_k),q(b_k),r(b_k))$, where $\pi:[1,n]\to[1,n]$ is again the permutation sending the first $n/2$ elements to the odd positions and the $n/2$ last ones to the even positions. Notice first that for every integer $k\in[1,n]$, it holds that $(p_k,q_k,r_k)\in\{(p\pm1,q,r\mp 1),(p,q\pm 1,r\mp 1)\}$. Furthermore, observe that if $k\in[0,n/2]$ and $(p_k,q_k,r_k)=(p+\alpha,q+\beta,r+\gamma)$ for $\{\alpha,\beta,\gamma\}=\{0,\pm 1\}$ with $\gamma\neq 0$, then we have $(p_{k+n/2},q_{k+n/2},r_{k+n/2})=(p+\alpha',q+\beta',r+\gamma')$ for some $\{\alpha',\beta',\gamma'\}=\{0,\pm 1\}$ with $\gamma'\neq 0$, where $\alpha'$,$\beta'$ and $\gamma'$ only depend on $\alpha$, $\beta$ and $\gamma$. Explicitly, we have $(\alpha',\beta',\gamma')=(\alpha,\beta,\gamma)$ if $\gamma=1$ and $(\alpha',\beta',\gamma')=(\beta,\alpha,\gamma)$ if $\gamma=-1$.

@@

In Table~\ref{table:comparisonsDegAI}, we also display the algebraic degree of different weightwise quadratic functions. We observe that the revisited \hwbf{} is the function with the highest degree, together with the \hwbf{}.

@@

\bibliographystyle{splncs04}
\bibliography{add}
